---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


### **load all packages used in this exploratory analysis**

```{r, echo=TRUE}

library(dplyr)
library(ggplot2)

```

### **Loading and preprocessing the data**

##### First, we read the data in from the zipped csv file and omit NAs:

```{r, echo=TRUE}

unzip('activity.zip')
activityData <- read.csv('activity.csv')

activityData_WithoutNan <- activityData[ with (activityData, { !(is.na(steps)) } ), ]

```

### **What is mean total number of steps taken per day?**

##### Let's Calculate the total number of steps taken per day :

```{r, echo=TRUE}

by_day <- group_by(activityData_WithoutNan, date)
steps_by_day <- summarise(by_day, total = sum(steps))
steps_by_day


```

##### create a histogram of the total number of steps per day : 

```{r, echo=TRUE}

hist(steps_by_day$total, main="Histogram of total number of steps per day", xlab="Total number of steps in a day")

```

##### calculate the mean and median values : 

```{r, echo=TRUE}
stepsByDayMean <- mean(steps_by_day$total)
stepsByDayMedian <- median(steps_by_day$total)

```

### **What is the average daily activity pattern?**

##### Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r, echo=TRUE}

steps_by_interval <- aggregate(steps ~ interval, activityData_WithoutNan, mean)

plot(steps_by_interval$interval, steps_by_interval$steps, type='l', 
     main="Average number of steps over all days", xlab="Interval", 
     ylab="Average number of steps")


```

##### Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=TRUE}

# find row with max of steps
max_steps_row <- which.max(steps_by_interval$steps)

# find interval with this max
steps_by_interval[max_steps_row, ]

```

### **Imputing missing values**

##### Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r, echo=TRUE}

sum(is.na(activityData))

```
 
##### My strategy is replacing NA’s with the mean for that 5-minute interval

```{r, echo=TRUE}

data_imputed <- activityData
for (i in 1:nrow(data_imputed)) {
  if (is.na(data_imputed$steps[i])) {
    interval_value <- data_imputed$interval[i]
    steps_value <- steps_by_interval[
      steps_by_interval$interval == interval_value,]
    data_imputed$steps[i] <- steps_value$steps
  }
}


```

##### Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r, echo=TRUE}

df_imputed_steps_by_day <- aggregate(steps ~ date, data_imputed, sum)
head(df_imputed_steps_by_day)


```

##### Make a histogram of the total number of steps taken each day

```{r, echo=TRUE}

hist(df_imputed_steps_by_day$steps, main="Histogram of total number of steps per day (imputed)", 
     xlab="Total number of steps in a day")

```

##### get mean and median of imputed data
```{r, echo=TRUE}

mean(df_imputed_steps_by_day$steps)

```

```{r, echo=TRUE}

median(df_imputed_steps_by_day$steps)

```

##### get mean and median of data without NA's
```{r, echo=TRUE}

mean(steps_by_day$total)

```

```{r, echo=TRUE}

median(steps_by_day$total)

```

##### **Mean values stays the same but therer is slight difference in meadian value**

### **Are there differences in activity patterns between weekdays and weekends?**

##### Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.

```{r, echo=TRUE}

data_imputed['type_of_day'] <- weekdays(as.Date(data_imputed$date))
data_imputed$type_of_day[data_imputed$type_of_day  %in% c('Saturday','Sunday') ] <- "weekend"
data_imputed$type_of_day[data_imputed$type_of_day != "weekend"] <- "weekday"

# convert type_of_day from character to factor
data_imputed$type_of_day <- as.factor(data_imputed$type_of_day)

```

##### plot a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis).

```{r, echo=TRUE}
# calculate average steps by interval across all days
df_imputed_steps_by_interval <- aggregate(steps ~ interval + type_of_day, data_imputed, mean)

# creat a plot
qplot(interval, 
      steps, 
      data = df_imputed_steps_by_interval, 
      type = 'l', 
      geom=c("line"),
      xlab = "Interval", 
      ylab = "Number of steps", 
      main = "") +
  facet_wrap(~ type_of_day, ncol = 1)

```